---
title: "Class06 extension"
author: "Michael Preston (PID: A53310268)"
format: pdf
---

# Section 1: Improving analysis code by writing functions

## A. 
Improve this regular R code by abstracting the main activities in your own new
function. Note, we will go through this example together in the formal lecture. The main
steps should entail running through the code to see if it works, simplifying to a core
working code snippet, reducing any calculation duplication, and finally transferring your
new streamlined code into a more useful function for you.

```{r}
df <- data.frame(a=1:10, b=seq(200,400,length=10),c=11:20,d=NA)
df
```

```{r}
norm <- function(data){
  nord_data <- (data - min(data)) / (max(data) - min(data))
  return(nord_data)
}

df[] <- lapply(df, norm)
df
```

## B. 
Next improve the below example code for the analysis of protein drug interactions by
abstracting the main activities in your own new function. Then answer questions 1 to 6
below. It is recommended that you start a new Project in RStudio in a new directory and
then install the bio3d package noted in the R code below (N.B. you can use the
command install.packages("bio3d") or the RStudio interface to do this).
Then run through the code to see if it works, fix any copy/paste errors before simplifying
to a core working code snippet, reducing any calculation duplication, and finally
transferring it into a more useful function for you.

```{r}
# install the bio3d package 
# install.packages("bio3d")
```

```{r}
# Can you improve this analysis code?
library(bio3d)

run_analysis <- function(pdb_name){
  s <- read.pdb(pdb_name)
  s.chainA <- trim.pdb(s, chain="A", elety="CA")
  s.b <- s.chainA$atom$b
  plotb3(s.b, sse=s.chainA, typ="l", ylab="Bfactor")
  return(s.b)
}

#for (pdb_name in c("4AKE", "1AKE", "1E4Y")){
#  run_analysis(pdb_name)
#}
s1.b <- run_analysis("4AKE")
s2.b <- run_analysis("1AKE")
s3.b <- run_analysis("1E4Y")
```

### Q1. 
What type of object is returned from the read.pdb() function?
```{r}
s <- read.pdb("4AKE")
typeof(s)
```
list

### Q2. 
What does the trim.pdb() function do?

```{r}
?trim.pdb()
```
trim.pdb() produces a new smaller PDB object, containing a subset of atoms. Here, trimmming to chain A.

### Q3. 
What input parameter would turn off the marginal black and grey rectangles in the
plots and what do they represent in this case?

```{r}
?plotb3

```
sse. they represent where the protein's secondary structures elements occur.

### Q4. 
What would be a better plot to compare across the different proteins?

A cluster dendrogram.

### Q5. 
Which proteins are more similar to each other in their B-factor trends. How could
you quantify this? HINT: try the rbind(), dist() and hclust() functions together with a
resulting dendrogram plot. Look up the documentation to see what each of these
functions does.

```{r}
hc <- hclust( dist( rbind(s1.b, s2.b, s3.b) ) )
plot(hc)

```
S2 and S3 are more similar
