---
title: "class10"
author: "Michael Preston (PID: A53310268)"
format: pdf
---

## Class 10: Halloween Mini-Project
Exploratory Analysis of Halloween Candy

## 1. Importing candy data

```{r}
#candy = read.csv("candy-data.csv", row.names=1)
library(readr)
# read in data and set column names
candy = read_csv("candy-data.csv", show_col_types = FALSE)
candy <- as.data.frame(candy)
rownames(candy) <- candy[[1]]
candy <- candy[, -1]
head(candy)

# flextable allows for scrollable table in rendered doc
flextable::flextable(head(candy))

```

```{r, messgae=FALSE}
library(dplyr)
```

Alternative way to run functions with 'pipe' in dplyr. This can be more readable for multi-step operations

```{r}
candy |> select(winpercent) |> sum()

```

> Q1. How many different candy types are in this dataset?

There are `r nrow(candy) ` different candy types

> Q2. How many fruity candy types are in the dataset?

There are `r sum(candy$fruity)` fruity candy types


## 2. What is your favorate candy?


> Q3. What is your favorite candy in the dataset and what is it’s winpercent value?

Reese's Peanut Butter cup: `r  round(candy["Reese's Peanut Butter cup", ]$winpercent)`

> Q4. What is the winpercent value for “Kit Kat”?

`r round(candy["Kit Kat", ]$winpercent)`

> Q5. What is the winpercent value for “Tootsie Roll Snack Bars”?

`r round(candy["Tootsie Roll Snack Bars", ]$winpercent) `

try out skimr package
```{r}
library("skimr")
skim(candy)
```

>Q6. Is there any variable/column that looks to be on a different scale to the majority of the other columns in the dataset?

winpercent

>Q7. What do you think a zero and one represent for the candy$chocolate column?

Logical/booleen indicator of whether the candy contains chocolate.

>Q8. Plot a histogram of winpercent values

```{r}
library("ggplot2")
ggplot(candy, aes(x = winpercent)) +
  geom_histogram(aes(y = after_stat(density))) +
  geom_density()

```

>Q9. Is the distribution of winpercent values symmetrical?

Somewhat right skewed

>Q10. Is the center of the distribution above or below 50%?

The median is below, while the mean is slightly above. 

mean: `r mean(candy$winpercent)`
median: `r median(candy$winpercent)`

>Q11. On average is chocolate candy higher or lower ranked than fruit candy?

Higher

```{r}
mean(candy[candy$fruity==1,]$winpercent)
mean(candy[candy$chocolate==1,]$winpercent)
```

>Q12. Is this difference statistically significant?

Yes

```{r}
t.test(candy[candy$fruity==1,]$winpercent, candy[candy$chocolate==1,]$winpercent)
```

## 3. Overall Candy Rankings


> Q13. What are the five least liked candy types in this set?

Nik L Nip, Boston Baked Beans, Chiclets, Super Bubble, and Jawbusters

```{r}
candy |> arrange(winpercent) |> head(5)

```

> Q14. What are the top 5 all time favorite candy types out of this set?

Reese's Peanut Butter cup, Reese's Miniatures, Twix, Kit Kat, and Snickers.

```{r}
candy |> arrange(winpercent) |> tail(5)
```

> Q15. Make a first barplot of candy ranking based on winpercent values.

```{r}
ggplot(candy) + 
  aes(winpercent, reorder(rownames(candy), winpercent)) + 
  geom_col() +
  ylab('Candy name') +
  xlab('Win percentage')
```

#### Time to add some useful color

```{r}
my_cols=rep("black", nrow(candy))
my_cols[as.logical(candy$chocolate)] = "chocolate"
my_cols[as.logical(candy$bar)] = "brown"
my_cols[as.logical(candy$fruity)] = "pink"

ggplot(candy) + 
  aes(winpercent, reorder(rownames(candy), winpercent)) + 
  geom_col(fill=my_cols) +
  ylab('Candy name') +
  xlab('Win percentage')

```

>Q17. What is the worst ranked chocolate candy?

`r rownames(arrange(candy[as.logical(candy$chocolate),], winpercent))[1] `

>Q18. What is the best ranked fruity candy?

Starburst


```{r}
temp <- arrange(candy[as.logical(candy$fruity),], winpercent)
rownames(temp)[nrow(temp)]
```

## 4. Taking a look at pricepercent

```{r, fig.height=6, fig.width=6}
library(ggrepel)

# How about a plot of price vs win
ggplot(candy) +
  aes(winpercent, pricepercent, label=rownames(candy)) +
  geom_point(col=my_cols) + 
  geom_text_repel(col=my_cols, size=3.3, max.overlaps = 20)
```

>Q19. Which candy type is the highest ranked in terms of winpercent for the least money - i.e. offers the most bang for your buck?

Reeses Miniatures is a good balance

>Q20. What are the top 5 most expensive candy types in the dataset and of these which is the least popular?

5 most expensive candy types: "Nik L Nip", "Nestle Smarties", "Ring pop", "Hershey's Krackel", "Hershey's Milk Chocolate"

Least popular: Nik L Nip

```{r}
ord <- order(candy$pricepercent, decreasing = TRUE)
head( candy[ord,], n=5 )
```
> Q21. Make a barplot again with geom_col() this time using pricepercent and then improve this step by step, first ordering the x-axis by value and finally making a so called “dot chat” or “lollipop” chart by swapping geom_col() for geom_point() + geom_segment()

```{r, fig.height=12, fig.width=6}
ggplot(candy) + 
  aes(pricepercent, reorder(rownames(candy), pricepercent)) + 
  geom_col() +
  ylab('Candy name') +
  xlab('Price percentage')
```

```{r, fig.height=12, fig.width=6}
# Make a lollipop chart of pricepercent
ggplot(candy) +
  aes(pricepercent, reorder(rownames(candy), pricepercent)) +
  geom_segment(aes(yend = reorder(rownames(candy), pricepercent), 
                   xend = 0), col=my_cols) +
  geom_point()
```

## 5 Exploring the correlation structure

```{r}
library(corrplot)
```

```{r, fig.height=6, fig.width=6}
cij <- cor(candy)
corrplot(cij)
```

> Q22. Examining this plot what two variables are anti-correlated (i.e. have minus values)?

Chocolate and Fruity have the greatest negative correlation.

> Q23. Similarly, what two variables are most positively correlated?

winpercent and chocolate

## 6. Principal Component Analysis

```{r}
# Define plotting functions for PCA results

plot_pca_scatter12 <- function(pca){
  # Plot "PC Plot" / "score plot" i.e. scatterplot of PC1 v. PC2
  ggplot(pca$x) +
    aes(x = PC1, y = PC2) +
    geom_point()
}

plot_pca_scree <- function(pca){
  # Create scree plot

  # Compute variance explained
  v <- round(pca$sdev^2/sum(pca$sdev^2) * 100)
  variance_df <- data.frame(
    PC = factor(paste0("PC", 1:length(v)), levels = paste0("PC", 1:length(v))),
    Variance = v
  )
  
  # Plot
  ggplot(variance_df) +
  aes(x = PC, y = Variance) +
  geom_col(fill = "steelblue") +
  xlab("Principal component") +
  ylab("Percent variation") +
  theme_bw() +
  theme(axis.text.x = element_text(angle = 0))
}

plot_pca_loading <- function(pca){
  # plot PCA loadings
  ggplot(pca$rotation) +
  aes(x = PC1,
  y = reorder(rownames(pca$rotation), PC1)) +
  geom_col(fill = "steelblue") +
  xlab("PC1 Loading Score") +
  ylab("") +
  theme_bw() +
  theme(axis.text.y = element_text(size = 9))
}

plot_pca_results <- function(pca){
  print(plot_pca_scatter12(pca))
  print(plot_pca_scree(pca))
  print(plot_pca_loading(pca))
}

```

```{r}
# Perform PCA and show results
candy.pca <- prcomp(candy, scale=T)
summary(candy.pca)
plot_pca_results(candy.pca)
```

```{r, fig.height=4, fig.width=6}
# repeat, plotting size based on winpercent

# Make a new data-frame with our PCA results and candy data
my_data <- cbind(candy, candy.pca$x[,1:3])

# plot
p <- ggplot(my_data) + 
        aes(x=PC1, y=PC2, 
            size=winpercent/100,  
            text=rownames(my_data),
            label=rownames(my_data)) +
        geom_point(col=my_cols)

p
```

Use ggrepel package and the function ggrepel::geom_text_repel() to label up the plot with non overlapping candy names 
```{r, fig.height=6, fig.width=6}
p + geom_text_repel(size=3.3, col=my_cols, max.overlaps = 30)  + 
  theme(legend.position = "none") +
  labs(title="Halloween Candy PCA Space",
       subtitle="Colored by type: chocolate bar (dark brown), chocolate other (light brown), fruity (red), other (black)",
       caption="Data from 538")
```

One more plotting option with gploty. This is an interactive plot, so you can get info on each data point by hovering the mouse over it
```{r}
# library(plotly)
# ggplotly(p)
```

Check out loadings
```{r}
par(mar=c(8,4,2,2))
barplot(candy.pca$rotation[,1], las=2, ylab="PC1 Contribution")
```

> Q24. What original variables are picked up strongly by PC1 in the positive direction? Do these make sense to you?

Fruity, hard, and pluribus. Yes, these features describe the candies with the highest values along PC1 (e.g., skittles, warheads, lemonheads)